MOMS-PI: Key dynamics

### 2022-01-11 notebook/_posts/2021-11-01-momspi-summary/momspi-summary.Rmd
Revise MOMSPI key-dynamics analysis
- Replace zeros prior to calibration and downstream analysis
- Create plots of example community trajectories
- Perform additional checks and calculations

@@

Note, I'm not using pseudo-props so the zeros are staying zero.
TODO: Reconsider this choice.

@@

@@

# (TODO) Example community trajectory

Pick a woman with multiple visits and a large change in mean efficiency

```{r}
tb %>%
  filter(host_subject_id.1 == host_subject_id.2) %>%
  mutate(fc = mean_efficiency.2 / mean_efficiency.1) %>%
  arrange(desc(abs(log(fc)))) %>%
  select(fc, host_subject_id.1, top_plot.1, top_plot.2, host_visit_rank.1)
```

# Testing

```{r}
cts <- momspi %>% sample_data %>% as_tibble %>% count(host_subject_id)
nrow(cts)
#> [1] 577
cts %>% pull(n) %>% summary
#>    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#>    1.00    2.00    3.00    3.39    4.00    9.00 
cts %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 1)
  # geom_dotplot(binwidth = 1)
```

577 subjects

>>>>

Finally, let's impute zeros using the Dirichlet posterior mean, with the prior respecting the average proportions across samples.

```{r}
taxa_mean_prop <- momspi %>%
  otu_table %>%
  transform_sample_counts(close_elts) %>%
  orient_taxa(as = 'cols') %>%
  as('matrix') %>%
  colMeans
stopifnot(sum(taxa_mean_prop) == 1)
prior_vec <- taxa_mean_prop * ntaxa(momspi)
stopifnot(identical(length(prior_vec), ntaxa(momspi)))
# Note the need for the seq_along trick to get phyloseq to allow this
# adjustment
momspi <- momspi %>% transform_sample_counts(~ prior_vec[seq_along(.x)] + .x)
```

## Checks

```{r}
cts <- momspi %>% sample_data %>% as_tibble %>% count(host_subject_id)
nrow(cts)
```

how does the number of visits vary by subject?

```{r}
cts %>% pull(n) %>% summary
cts %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 1)
```

@@

@@

# Variation in mean efficiency and individual species

```{r}
focal_taxa <- momspi %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
  taxa_names %>%
  print
```

```{r}
x <- momspi_cal %>%
  transform_sample_counts(close_elts) %>%
  filter_tax_table(.otu %in% focal_taxa) %>%
  as_tibble
```

```{r}
taxa_summary <- x %>%
  mutate(across(.abundance, log)) %>%
  with_groups(.otu, summarize,
    across(.abundance, 
      .fns = c(mean = mean, median = median, sd = sd, var = var),
      .names = '{.fn}'
    )
  )
taxa_summary %>% arrange(-mean) %>% print(n=Inf)
```

```{r}
me_summary <- mean_eff1 %>%
  mutate(across(mean_efficiency, log)) %>%
  summarize(
    across(mean_efficiency,
      .fns = c(sd = sd, var = var),
      .names = '{.fn}'
    )
  ) %>%
  print
```

The mean efficiency has a smaller dispersion than individual species, but it is of a similar magnitude, so there is still potential for it to have a substantial impact.

What about variation within women?

```{r}
taxa_summary_within <- x %>%
  mutate(across(.abundance, log)) %>%
  with_groups(c(.otu, host_subject_id), summarize,
    across(.abundance,
      .fns = c(mean = mean, median = median, sd = sd, var = var),
      .names = '{.fn}'
    ),
    n = n()
  ) %>%
  filter(n >= 3) %>%
  with_groups(.otu, summarize,
    sd_mean = sd(mean),
    mean_sd = mean(sd)
  )
taxa_summary_within %>% print(n=Inf)
```

```{r}
me_summary_within <- mean_eff1 %>%
  mutate(across(mean_efficiency, log)) %>%
  with_groups(host_subject_id, summarize,
    across(mean_efficiency,
      .fns = c(mean = mean, sd = sd, var = var),
      .names = '{.fn}'
    ),
    n = n()
  ) %>%
  filter(n >= 3) %>%
  summarize(
    sd_mean = sd(mean),
    mean_sd = mean(sd)
  ) %>%
  print
```

Note, a better way to do this would perhaps be to use anova / linear modeling, or whatever method from those would properly weight the number of samples from each woman.

what about the variation between the first samples from each woman?

```{r}
taxa_summary_first <- x %>%
  with_groups(host_subject_id, slice_min, host_visit_number, n = 1) %>%
  mutate(across(.abundance, log)) %>%
  with_groups(.otu, summarize,
    across(.abundance,
      .fns = c(mean = mean, median = median, sd = sd, var = var),
      .names = '{.fn}'
    )
  )
taxa_summary_first %>% arrange(-mean) %>% print(n=Inf)
# taxa_summary %>% arrange(-mean) %>% print(n=Inf)
```

These sd's are similar to the overall sds, which makes sense given most pairwise comparisons are between women.

```{r}
me_summary_first <- mean_eff1 %>%
  with_groups(host_subject_id, slice_min, host_visit_number, n = 1) %>%
  mutate(across(mean_efficiency, log)) %>%
  summarize(
    across(mean_efficiency,
      .fns = c(sd = sd, var = var),
      .names = '{.fn}'
    )
  ) %>%
  print
me_summary
```

This is somewhat larger, but again similar.

Tentative: 
We see that the (log-e) mean efficiency sd is around 0.9, while the sd of individual taxa ranges from 2-3.5 (between women and overall).
Within women, the mean efficiency sd and that of individual taxa are both smaller, such that the relative importance of each seems similar.

Visualize, for the first sample from each woman:

```{r}
bind_rows(
  x, 
  mean_eff1 %>% mutate(.otu = 'Mean efficiency', .abundance = mean_efficiency)
) %>%
  with_groups(c(.otu, host_subject_id), slice_min, host_visit_number, n = 1) %>%
  mutate(
    across(.abundance, log),
    across(.otu, fct_reorder, .abundance)
  ) %>%
  ggplot(aes(y = .otu, x = .abundance)) +
  ggdist::stat_slab()
```

@@

# Example community trajectory

Pick a woman with multiple visits and a large change in mean efficiency to illustrate the potential impact of bias on species trajectories.
We expect large changes in efficiency to be drive by changes in dominance from Gard to Lactobacillus (or vice versa).

To pick a woman, let's first examine the distribution of geometric standard deviations of mean efficiency across subjects, and pick a subject at the upper end who has multiple visits.

```{r}
me <- mean_eff1 %>%
  mutate(across(mean_efficiency, log)) %>%
  with_groups(host_subject_id, summarize,
    across(mean_efficiency,
      .fns = c(mean = mean, sd = sd, var = var),
      .names = '{.fn}'
    ),
    n = n()
  ) %>%
  filter(n >= 3)
```

```{r}
me %>% slice_max(sd, n = 10)
```

Let's pick subject 'EP280575', who has the most visits within the top 10.

```{r}
sid <- 'EP280575'
ps_obs <- momspi %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~max(.) > 1e-3)
ps_cal <- momspi_cal %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  prune_taxa(taxa_names(ps_obs), .)
x <- list(Calibrated = ps_cal, Observed = ps_obs) %>%
  map_dfr(as_tibble, .id = "type") %>%
  mutate(across(.otu, fct_relabel, function(x) str_replace_all(x, '_', ' ')))
```

```{r, fig.dim = c(6, 15)}
taxa_to_plot <- x %>% 
  filter(type == 'Calibrated') %>%
  with_groups(.otu, summarize, 
    med = median(log10(.abundance)),
    max = max(log10(.abundance)),
  ) %>%
  filter(max > -2.5) %>%
  arrange(-max) %>%
  pull(.otu) %>%
  print
x %>%
  filter(.otu %in% taxa_to_plot) %>%
  mutate(.otu = factor(.otu, taxa_to_plot)) %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
  geom_line(aes(group = type)) +
  labs(y = 'Proportion', title = 'Species proportions') +
  scale_color_brewer(type = 'qual', palette = 6) +
  theme(legend.position = 'top')
```

For the final plot, I'll pick the most abundant 3 taxa (which are driving the mean efficiency) and two more taxa that show the different behaviors.

```{r trajectories-within-subject-high-var, fig.dim = c(6, 9)}
taxa_to_plot <- c(
  'Gardnerella_vaginalis', 'Lachnospiraceae_BVAB1', 'Lactobacillus_iners',
  'Ureaplasma_cluster23', 'Megasphaera_OTU70_type1') %>%
  str_replace_all('_', ' ')
p_sp <- x %>%
  filter(.otu %in% taxa_to_plot) %>%
  mutate(.otu = factor(.otu, taxa_to_plot)) %>%
  filter(as.integer(.otu) <= 6) %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
  geom_line(aes(group = type)) +
  labs(y = 'Proportion', title = 'Species proportions') +
  scale_color_brewer(type = 'qual', palette = 6) +
  theme(legend.position = 'top')
p_me <- mean_eff1 %>%
  filter(host_subject_id == sid) %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  scale_y_log10() +
  labs(y = 'Efficiency', title = 'Mean efficiency') +
  geom_line(aes(group = 1))
p_me / p_sp &
  annotate('rect', xmin = 3, xmax = 4, ymin = Inf, ymax = 0, alpha = 0.1) &
  labs(x = 'Host visit') &
  plot_layout(heights = c(0.1, 1)) &
  plot_annotation(tag_levels = 'A')
```

For comparison, let's also look at a long trajectory where there is no major change in mean efficiency.

```{r}
me %>% slice_min(sd, n = 10)
```

```{r}
sid <- 'EP694646'

ps_obs <- momspi %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~max(.) > 1e-3)
ps_cal <- momspi_cal %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  prune_taxa(taxa_names(ps_obs), .)
x <- list(Calibrated = ps_cal, Observed = ps_obs) %>%
  map_dfr(as_tibble, .id = "type") %>%
  mutate(across(.otu, fct_relabel, function(x) str_replace_all(x, '_', ' ')))
ntaxa(ps_obs)
```

```{r trajectories-within-subject-low-var, fig.dim = c(6, 9)}
p_sp <- x %>%
  mutate(.otu = fct_reorder(.otu, -.abundance)) %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
  geom_line(aes(group = type)) +
  labs(y = 'Proportion', title = 'Species proportions') +
  scale_color_brewer(type = 'qual', palette = 6) +
  theme(legend.position = 'top')
p_me <- mean_eff1 %>%
  filter(host_subject_id == sid) %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  scale_y_log10() +
  labs(y = 'Efficiency', title = 'Mean efficiency') +
  geom_line(aes(group = 1))
p_me / p_sp &
  labs(x = 'Host visit') &
  plot_layout(heights = c(0.1, 1)) &
  plot_annotation(tag_levels = 'A')
```

Note that the variation in mean efficiency is very small.
